---
title: "SVM Model for Predicting Soccer Match Winners"
author: "Ismayil, Vadym, Ivan"
date: "`r Sys.Date()`"
output: html_document
---

# Project Description
- This project is aimed at using machine learning to predict the outcomes of football matches. More specifically, it uses the Support Vector Machines (SVM) model to determine the winning team. The project has been jointly undertaken by Ismayil, Vadym, and Ivan.
- This project was undertaken as a part of an academic task that required the translation of statistical analysis from one programming language to another. The original study, conducted in Python, was translated into R, and the analysis was reproduced. The process offered us an opportunity to understand the nuances of different programming languages and how the same task can be executed differently in each.

## Data Acquisition and Preparation
- The data for this project comes from two CSV files: matches_1930_2022.csv and world_cup.csv. The first file contains data on soccer matches from 1930 through 2022, while the second one provides information on World Cup matches.
- In the data preparation phase, unneeded columns from the matches dataset are dropped, focusing on essential match details such as home team, away team, and scores. Teams' names are replaced with appropriate names where necessary. A numeric index is assigned to each unique team name for more manageable data processing.
- Next, the championships won by each team are counted, and new columns, Home_Team_Champions and Away_Team_Champions, are added to the matches dataset, indicating how many times the home and away teams have won championships, respectively.
- The winning team for each match is calculated and added to the dataset as the column Winner.

## Data Analysis
- After data preparation, the dataset is visualized to see the distribution of World Cup championships by country. Next, teams' names are replaced with their corresponding numeric indices. Finally, missing values are replaced with appropriate values.
- The dataset is then split into training and test sets, with 70% of the data used for training the SVM model and 30% used for testing it.

## Modeling
- A Support Vector Machines (SVM) model is used to predict match outcomes. The model is trained on the Winner attribute with other columns as features.

## Model Evaluation
- The model's predictions on both training and test data are computed, and the accuracy of the model is determined by comparing the model's predictions to the actual results. A confusion matrix is generated using the caret package's confusionMatrix() function to provide a detailed evaluation of the model's performance.
- The model's performance is visualized, and a comparison between the predicted and actual results is made. This helps in understanding how well the model has performed in predicting the outcomes of the matches.
- In conclusion, the project demonstrates how machine learning can be applied in the domain of sports analytics to predict soccer match outcomes using SVM.

```{r}
install.packages("e1071")
library(e1071)
library(caret)
library(dplyr)
library(tidyverse)
library(ggplot2)
```


```{r}
# Load Data
matches_1930_2022 <- read.csv("matches_1930_2022.csv")
```


```{r}
world_cup <- read.csv("C:/Users/Vadym/Desktop/Reproducible/project/world_cup.csv")
```


```{r}
# Data Preparation
matches <- matches_1930_2022[, -c(9, 10, 11, 12, 14,15,18,19,20,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,43,44,45)]
```


```{r}
replace_teams <- function(matches) {
  matches$home_team[matches$home_team %in% c("German DR", "Germany FR")] <- "Germany"
  matches$home_team[matches$home_team == "Soviet Union"] <- "Russia"
  
  matches$away_team[matches$away_team %in% c("German DR", "Germany FR")] <- "Germany"
  matches$away_team[matches$away_team == "Soviet Union"] <- "Russia"
  
  return(matches)
}

#### Activate function replace_teams
matches <- replace_teams(matches)
```


```{r}
team_name <- list()
index <- 0

for (i in 1:nrow(matches)) {
  name <- matches[i, "home_team"]
  if (!(name %in% names(team_name))) {
    team_name[[name]] <- index
    index <- index + 1
  }
  
  name <- matches[i, "away_team"]
  if (!(name %in% names(team_name))) {
    team_name[[name]] <- index
    index <- index + 1
  }
}

team_name
```



```{r}
championships <- table(ifelse(world_cup$Champion == "West Germany", "Germany", world_cup$Champion))
championships
```



```{r}
# Create a countplot
ggplot(world_cup, aes(x = Champion)) +
  geom_bar(fill = "steelblue") +
  labs(x = "Country", y = "Number of World Cup Championships") +
  ggtitle("Number of World Cup Championships by Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




```{r}
matches$Home_Team_Champions <- 0
matches$Away_Team_Champions <- 0


if (!is.null(championships[matches$home_team])) {
  matches$Home_Team_Champions <- championships[matches$home_team]
}

if (!is.null(championships[matches$away_team])) {
  matches$Away_Team_Champions <- championships[matches$away_team]
}

```


```{r}
matches$Winner <- "-"

calculate_winner <- function(matches) {
  home_team_goal <- as.integer(matches$home_score)
  away_team_goal <- as.integer(matches$away_score)
  matches$Winner <- ifelse(home_team_goal == away_team_goal, 0, ifelse(home_team_goal > away_team_goal, 1, 2))
  
  return(matches)
}

matches <- calculate_winner(matches)

table(matches$Winner)
```


```{r}
replace_team_name <- function(matches) {
  matches$home_team <- team_name[matches$home_team]
  matches$away_team <- team_name[matches$away_team]
  return(matches)
}

matches <- replace_team_name(matches)
```


```{r}
matches <- matches[, -c(5, 8, 9, 10, 11, 12, 13)]
matches <- matches[, -c(3,4,5,6)]


matches$Away_Team_Champions <- ifelse(is.na(matches$Away_Team_Champions), 0, matches$Away_Team_Champions)
matches$Home_Team_Champions <- ifelse(is.na(matches$Home_Team_Champions), 0, matches$Home_Team_Champions)

```

```{r}
# Unlist the dataset
class(matches)

matches$Home_Team_Champions<-as.numeric(matches$Home_Team_Champions)
matches$Away_Team_Champions<-as.numeric(matches$Away_Team_Champions)
matches$away_team <- unlist(matches$away_team)
matches$home_team <- unlist(matches$home_team)

#mean_value <- mean(matches$home_xg, na.rm = TRUE)
#matches$home_xg <- ifelse(is.na(matches$home_xg), mean_value, matches$home_xg)
#mean_value1 <- mean(matches$away_xg, na.rm = TRUE)
#matches$away_xg <- ifelse(is.na(matches$away_xg), mean_value1, matches$away_xg)
```

```{r}
# Create indices for the training and test sets
sample <- sample(c(TRUE, FALSE), nrow(matches), replace=TRUE, prob=c(0.7,0.3))
train <- matches[sample, ]
test <- matches[!sample, ]  
```

```{r}
####modeling 

model <- svm(Winner ~.,
                 data = train,
                 type = 'C-classification',
                 kernel = 'linear',
                scale = FALSE)
model
```



```{r}
pred_test <- predict(model,newdata = test)
pred_test
```

```{r}
pred_train1 <- predict(model, train)
mean(pred_train1 == train$Winner)
```


```{r}
####evaluation
pred_test <- predict(model, test)
mean(pred_test == test$Winner)
```


```{r}
# Create a data frame with the true and predicted labels
results <- data.frame(Actual = test$Winner, Predicted = pred_test)
```


```{r}
# Use the caret package's confusionMatrix() function to calculate evaluation metrics
cm <- confusionMatrix(data = factor(results$Predicted),
                      reference = factor(results$Actual))
# Print the classification report
print(cm)
```

